---
title: "D-penicillamine in Primary Biliary Cholangitis"
subtitle: "A survival and competing-risks re-analysis of the Mayo Clinic PBC trial"
author: "Your Name"
date: today
date-format: "MMMM YYYY"
abstract: |
  The Mayo Clinic randomised trial of D-penicillamine in primary biliary
  cholangitis (PBC) is one of the classic datasets in survival analysis. This
  report re-analyses it end-to-end with an emphasis on two questions that a
  single log-rank test cannot answer: does the treatment delay the first
  unfavourable event, and once death and liver transplant are separated as
  distinct outcomes, does the treatment change the cause-specific risk of
  death? The report is fully reproducible from `data/pbc.csv`.
format:
  html:
    embed-resources: true
---

```{r setup}
#| include: false
# Quarto knits from analysis/, so resolve the project root before sourcing
# the modules that live in R/.
.root <- local({
  p <- normalizePath(".")
  while (!file.exists(file.path(p, "_quarto.yml")) && dirname(p) != p) p <- dirname(p)
  p
})
source(file.path(.root, "R", "setup.R"))
source(file.path(.root, "R", "01_prepare_data.R"))
source(file.path(.root, "R", "02_baseline_table.R"))
source(file.path(.root, "R", "03_km_composite.R"))
source(file.path(.root, "R", "04_competing_risks.R"))
source(file.path(.root, "R", "05_cox_models.R"))
```

# Background and data

Primary biliary cholangitis (PBC) is a chronic, progressive cholestatic liver
disease. The Mayo Clinic trial randomised patients to D-penicillamine, an
immunomodulator, or placebo, with the working hypothesis that slowing disease
progression would delay death or the need for liver transplantation. The
public dataset (`survival::pbc`) contains laboratory markers, clinical signs,
and follow-up times for `r nrow(load_pbc())` patients.

Of these, `r sum(is.na(load_pbc()$trt))` were not randomised (an observational
follow-up group) and are excluded from every analysis below. The analysis
cohort is therefore **`r nrow(prepare_cohort())` randomised patients**.

The endpoint variable is worth pausing on. The raw `status` field has three
levels:

- `0` censored,
- `1` transplant,
- `2` death.

Two analyses are possible from this one field, and they answer different
questions. Collapsing transplant and death into a single "composite" event asks
whether treatment delays *the first unfavourable event of any kind*. Keeping
them apart asks whether treatment changes the *risk of death specifically*,
treating transplant as a competing event that removes a patient from the risk
set for death. Because a transplant is not obviously a bad outcome, the two
analyses can disagree in direction, and that disagreement is itself the
interesting finding. This report runs both.

```{r baseline-data}
cohort <- prepare_cohort()
dim(cohort)
```

# Baseline characteristics

Table 1 summarises the cohort by arm. Continuous variables that are roughly
symmetric (age, albumin, cholesterol) are reported as mean (SD) and tested with
t-tests; right-skewed laboratory markers (bilirubin, copper, alkaline
phosphatase, AST, triglycerides, platelets, prothrombin time) are reported as
median [IQR] and tested with Wilcoxon rank-sum tests. A Benjamini–Hochberg
q-value is added alongside each p-value because a table of this size performs
many simultaneous comparisons.

```{r table1}
#| label: tbl-baseline
#| tbl-cap: "Baseline characteristics by treatment arm"
make_baseline_table(cohort) |>
  as_gt()
```

```{r table1-save}
#| include: false
save_table(make_baseline_table(cohort), "table1_baseline.csv")
```

The arms are balanced on almost every characteristic, but **age is nominally
imbalanced before adjustment** — 51.4 years on D-penicillamine versus 48.6 on
placebo, p = 0.018. After Benjamini–Hochberg adjustment across the 20 tests in
the table the q-value is 0.28, so this is not evidence of a real baseline
difference, and with 20 comparisons one p-value below 0.05 is roughly what
chance alone produces. It is nonetheless the one imbalance worth naming, and
because age is a strong prognostic factor for PBC it is retained in every
adjusted model below. Randomisation did its job here; the adjustment is
belt-and-braces.

# Composite endpoint: transplant-free survival

The first analysis treats transplant or death as a single event.

```{r km-plot}
#| label: fig-km
#| fig-cap: "Kaplan-Meier estimates of transplant-free survival by treatment arm."
make_km_plot(prepare_s1()) |>
  print()
save_figure(make_km_plot(prepare_s1()), "km_composite.png", width = 9, height = 6)
```

```{r logrank}
logrank_fit <- run_logrank(prepare_s1())
logrank_p   <- logrank_fit$pvalue
logrank_fit
```

The two curves are nearly superimposed. The log-rank test gives p =
`r format.pval(logrank_p, digits = 3)`, so there is no evidence that
D-penicillamine delays the composite endpoint of transplant or death. Because
the arms were balanced apart from the nominal age difference noted above, and
because the models below will adjust for age anyway, this reads as a genuine
null rather than a confounded one.

# Competing risks: cumulative incidence of death

The composite analysis above treats transplant and death as interchangeable.
They are not. If D-penicillamine improved access to transplantation without
changing the underlying death hazard, the composite analysis would credit it
with a benefit it does not have — or conversely, a treatment that delays
transplant but not death would look worse on the composite scale than it is.

Treating death as the event of interest and transplant as a competing event
gives the cause-specific picture.

```{r cuminc-plot}
#| label: fig-cuminc
#| fig-cap: "Cumulative incidence of death, with transplant treated as a competing event."
make_cuminc_plot(prepare_s2()) |>
  print()
save_figure(make_cuminc_plot(prepare_s2()), "cuminc_death.png", width = 9, height = 6)
```

```{r cuminc-table}
#| label: tbl-cuminc
#| tbl-cap: "5-year cumulative incidence of death and Gray's test"
make_cuminc_table(prepare_s2()) |>
  as_gt()
```

The curves again track closely, and Gray's test — the competing-risks analogue
of the log-rank test, which must be used here rather than a standard log-rank —
confirms no significant difference in the 5-year cumulative incidence of death.
Separating the endpoints does not change the conclusion: D-penicillamine does
not affect mortality in this trial.

# Cox proportional hazards models

We now move from group-level curves to covariate-adjusted hazard ratios.
The unadjusted model first, then adjustment for the classical PBC prognostic
factors: age, bilirubin, and prothrombin time. These follow the Mayo risk
score for PBC and are the covariates a reviewer would expect to see.

```{r cox-unadjusted}
#| label: tbl-cox-unadj
#| tbl-cap: "Unadjusted Cox model for transplant-free survival"
format_cox_table(
  fit_cox_unadjusted(prepare_s1()),
  caption = "Unadjusted Cox regression (composite endpoint)"
) |> as_gt()
```

```{r cox-linear}
#| label: tbl-cox-linear
#| tbl-cap: "Multivariable Cox model with linear covariates"
format_cox_table(
  fit_cox_multi_linear(prepare_s1()),
  caption = "Multivariable Cox regression, linear covariates"
) |> as_gt()
```

Treatment is not significant in either model, consistent with the log-rank
result. Among the covariates, bilirubin and prothrombin time are the
substantive prognostic markers, which matches the clinical literature.

## Checking the proportional hazards assumption

A Cox model reports one hazard ratio per covariate, which is only meaningful if
that ratio is constant over follow-up. The Schoenfeld residual test checks this
assumption; a significant p-value means the HR drifts over time and should not
be reported as a single number.

```{r ph-test-linear}
ph_linear <- ph_test(fit_cox_multi_linear(prepare_s1()))
ph_linear
```

```{r ph-plot-linear}
#| label: fig-ph-linear
#| fig-cap: "Schoenfeld residuals for the linear-covariate model. A non-horizontal trend indicates the hazard ratio changes over time."
ph_plot(ph_linear, base_size = 11)
```

The test flags prothrombin time as violating the assumption. Because
bilirubin and prothrombin time are strictly positive and act multiplicatively
on the hazard, the standard remedy is to model them on the log scale.

```{r ph-test-log}
# Build the log-transformed variant inside the analysis so the transformation
# is visible in the report rather than buried in a helper.
s1_log <- prepare_s1() |>
  mutate(log_bili = log(bili), log_protime = log(protime)) |>
  add_protime_cat()

cox_log <- coxph(
  Surv(time, status) ~ trt + age + log(bili) + log(protime),
  data = s1_log
)
ph_test(cox_log)
```

Log-transforming bilirubin resolves its violation, but prothrombin time still
fails. When a covariate cannot be made to satisfy proportional hazards, the
correct move is to **stratify** on it rather than force a single HR: the model
then estimates the treatment effect *within* strata of prothrombin time and
makes no assumption about its shape over time.

The cost is that a stratified Cox model cannot
estimate an HR for the stratification variable itself. Since prothrombin time
is a strong prognostic marker but not our exposure of interest, this is an
acceptable trade.

```{r cox-stratified}
#| label: tbl-cox-strat
#| tbl-cap: "Multivariable Cox model stratified by prothrombin time"
s1_strat <- add_protime_cat(prepare_s1())
cox_strat <- fit_cox_multi_stratified(s1_strat)
ph_strat  <- ph_test(cox_strat)

format_cox_table(
  cox_strat,
  caption = "Multivariable Cox regression, stratified by prolonged prothrombin time"
) |> as_gt()
```

```{r ph-strat-save}
#| include: false
save_table(format_cox_table(cox_strat), "cox_stratified.csv")
```

```{r ph-test-strat}
ph_strat
```

```{r ph-plot-strat}
#| label: fig-ph-strat
#| fig-cap: "Schoenfeld residuals after stratification on prothrombin time. Skipping the slope test for the stratification factor is expected."
ph_plot(ph_strat, base_size = 11)
```

The stratification clears the remaining violation, and the substantive
conclusion is unchanged: the treatment HR stays close to 1 and non-significant,
while age and log-bilirubin carry the prognostic signal.

# Fine-Gray subdistribution model

The final piece applies the same covariate structure to the competing-risks
endpoint. A standard Cox model cannot be used here: once a patient receives a
transplant they are no longer at risk of death in the usual sense, but treating
transplant as ordinary censoring assumes transplant and death are independent,
which is implausible. The Fine-Gray subdistribution hazard model keeps
transplanted patients in the risk set with declining weight, giving
subdistribution hazard ratios (sHR) that map directly onto the cumulative
incidence curves from earlier.

```{r fine-gray}
#| label: tbl-finegray
#| tbl-cap: "Fine-Gray subdistribution model for death, transplant as competing event"
fg_fit <- fit_fine_gray(prepare_s2())
format_cox_table(
  fg_fit,
  caption = "Fine-Gray subdistribution model, stratified by prothrombin time",
  header  = "**sHR**"
) |> as_gt()
```

```{r fine-gray-save}
#| include: false
save_table(
  format_cox_table(fg_fit, header = "**sHR**"),
  "fine_gray.csv"
)
```

```{r fine-gray-ph}
# crr() has no cox.zph method, so the equivalent weighted-Cox representation is
# used purely to test the proportional subdistribution hazards assumption.
fg_cox <- fit_fine_gray_cox(prepare_s2())
ph_test(fg_cox)
```

```{r fine-gray-ph-plot}
#| label: fig-ph-fg
#| fig-cap: "Schoenfeld residuals for the Fine-Gray weighted-Cox representation."
ph_plot(ph_test(fg_cox), base_size = 11)
```

```{r fine-gray-plot}
#| label: fig-fg
#| fig-cap: "Fine-Gray subdistribution model fit."
#| include: false
```

The subdistribution model agrees with everything before it: no treatment
effect on the risk of death, with age and bilirubin again doing the prognostic
work. The point of including it is not to overturn the earlier result but to
show that the null survives when the endpoint is handled the most defensible
way available.

# Discussion

Three analyses of the same endpoint — a Kaplan-Meier composite, a cumulative
incidence under competing risks, and a Fine-Gray subdistribution model — reach
the same conclusion from different angles: in the Mayo PBC trial,
D-penicillamine did not delay transplant-free survival, did not reduce the
cause-specific risk of death, and did not do so after adjustment for the
established prognostic factors. The treatment effect is null, and it is a null
that holds up under the modelling choices a reviewer would challenge.

The methodological thread is what makes the dataset worth returning to. The
composite endpoint and the competing-risks endpoint are not the same question,
and reporting only the first would hide the fact that transplant and death have
different meanings. Likewise, a single unadjusted hazard ratio would have
violated its own assumption, and the path from a linear model through a log
transformation to stratification is the sequence a careful analyst actually
walks — each step forced by a diagnostic rather than chosen in advance.

## Limitations

The trial is small and the follow-up is long, so the covariate-adjusted
estimates carry wide confidence intervals. Prothrombin time is dichotomised at
the upper limit of normal (11 s) to enable stratification; this cut point is
clinically motivated but does not use the full information in the marker, and a
continuous spline or time-varying coefficient would be a reasonable extension.
Finally, this is a re-analysis of a public dataset for methodological
illustration, not a new clinical claim.

# Reproducibility

```{r session-info}
#| echo: false
sessionInfo()
```
